Converging High-Throughput and High-Performance Computing: A Case Study

### 2017-09-08 nge.tex
Address reviewers' comments: please review/iterate

\ifreview
Comments to address in this section:
\begin{itemize}
	\color{red} 
	\item REVIEWER 3: The first paragraphs of section V. are basically a
	rediscovery of the advantages of pilot jobs for WLCG task scheduling.
	Since all those pain points were well-known in the WLCG community before
	the project started, it should be explained why a pilot job approach was
	not taken in the first place.
	\item REVIEWER 3: Figure 6. refers to a "User Workstation" as the running
	place for the Pilot and Unit Managers, which is never mentioned in the
	text. (I got the impression that the NGE would run on a DTN node; if this
	is not the case and separate workstations are being used for the testing
	phase, this should be made clear.)
\end{itemize}
\fi

@@

NGE exposes an API to describe workloads (Fig.~\ref{fig:arch-overview}, green
squares) and pilots (Fig.~\ref{fig:arch-overview}, red circles), and to
instantiate a PilotManager and a UnitManager. The PilotManager submits pilots
to Titan's PBS batch system via SAGA API (Fig.~\ref{fig:arch-overview}, dash
arrow). %, as done by PanDA Broker to submit MPI scripts. 
Once scheduled, the Pilot Agent is bootstrapped on a MOM node and the
Agent's Executors on the worker nodes.

The UnitManager and the Pilot Agent communicate via a database instantiated on a
DTN so as to be reachable by both modules. The UnitManager schedules units to
the Agent's Scheduler (Fig.~\ref{fig:arch-overview}, solid arrow) and the
Agent's Scheduler schedules the units on one or more Agent's Executor.
% for execution. The Agent's executors can manage one or more worker nodes,
% depending on performance evaluations.

The Pilot Agent uses the Open Run-Time Environment (ORTE) for communication and
coordination of the execution of units. This environment is a critical
component of the OpenMPI implementation~\cite{castain05:_open_rte}.
%, cug-2016}.
ORTE is able to minimize the system overhead while submitting tasks by
avoiding filesystem bottlenecks and race conditions with network sockets.

>>>>

\ifreview
Comments to address in this section:
\begin{itemize}
	\color{red} 
	\item REVIEWER 3: The first paragraphs of section V. are basically a
	rediscovery of the advantages of pilot jobs for WLCG task scheduling.
	Since all those pain points were well-known in the WLCG community before
	the project started, it should be explained why a pilot job approach was
	not taken in the first place.\mtnote{Added an opening paragraph recalling
	the explanation offered in \S\ref{sec:panda_titan}}.
	\item REVIEWER 3: Figure 6. refers to a "User Workstation" as the running
	place for the Pilot and Unit Managers, which is never mentioned in the
	text. (I got the impression that the NGE would run on a DTN node; if this
	is not the case and separate workstations are being used for the testing
	phase, this should be made clear.)\mtnote{Fixed.}
\end{itemize}
\fi

As explained in \S\ref{sec:panda_titan}, PanDA Broker was designed to
maximize code reutilization of PanDA Pilot. This allowed for rapid adoption
and incremental optimization while enabling the parallel development of a
more general solution for executing ATALS workloads on HPC resources.

@@

NGE exposes an API to describe workloads (Fig.~\ref{fig:arch-overview}, green
squares) and pilots (Fig.~\ref{fig:arch-overview}, red circles), and to
instantiate a PilotManager and a UnitManager. The PilotManager submits pilots
to Titan's PBS batch system via SAGA API (Fig.~\ref{fig:arch-overview}, dash
arrow). Once scheduled, the Pilot Agent is bootstrapped on a MOM node and the
Agent's Executors on the worker nodes.

The UnitManager and the Pilot Agent communicate via a database instantiated
on a DTN so as to be reachable by both modules. The UnitManager schedules
units to the Agent's Scheduler (Fig.~\ref{fig:arch-overview}, solid arrow)
and the Agent's Scheduler schedules the units on one or more Agent's
Executor.

The Pilot Agent uses the Open Run-Time Environment (ORTE) for communication
and coordination of the execution of units. This environment is a critical
component of the OpenMPI implementation~\cite{castain05:_open_rte}. ORTE is
able to minimize the system overhead while submitting tasks by avoiding
filesystem bottlenecks and race conditions with network sockets.

### 2017-09-08 overall.tex
Address reviewers' comments: please review/iterate

\ifreview
Comments to address in this section:
\begin{enumerate}
    \color{red} 
    \item REVIEWER 2: What are the reasons for the low efficiency of PanDA
    Broker? Why are only 30\% of backfill slots useable, if they are
    basically uncontested? Why does Fig 3 show an increase of absolute and
    relative efficiency in Aug-16, when the text claims this in Sep-16 due to
    more PanDA Brokers? Why are only 20 PanDA Brokers deployed if this is a
    limiting factor? Why is it questionable that more PanDA Brokers would
    increase efficiency? If only 5/6 of resources below the mean are
    unusable, this implies an efficiency of almost 60\% being reachable.
    \item REVIEWER 2: Given an expected increase in resource need of 10-100,
    how is the capacity of 4\% WLCG significant? How does this relate to the
    manpower required to develop and maintain a custom solution per HPC site?
    How many TITAN resources are actually available to HEP, given the use by
    other domains? The magnitude of capacities presented as available is in
    major contrast to the requirements sketched in the introduction.
    \item REVIEWER 2: How does the poor metadata read performance of
    1500-6300 s (25\%-100\% of minimum walltime!) affect efficiency? Is it
    factored into the choice of walltime requests? Is it applicable only to
    TITAN, to HPC in general, or even to regular WLCG sites? Are there
    solutions at application, e.g. the use of packaged python scripts and
    libraries? The information presented is insufficient to estimate whether
    this issue applies to other environments and usecases as well, or is
    simply due to suboptimal implementation/configuration.
    \item REVIEWER 3: The HEP-SPEC06 benchmark, which is used as a
    ``computing power'' measure in LCG is mostly integer-artihmetic based: FP
    operations accounts for \~10\% of the average workload (see [1], section
    ``Why do you use `all\_cpp' instead of SPECint or SPECfp?'') While this
    figure certainly comes from averaging over many different workloads
    across all experiments, it seems a bit surprising that FP scheduling
    contention is responsible for 50\% decrease in compute efficiency on
    Titan when compared to other WLCG sites (paper page 6).
\end{enumerate}
\fi

@@

% without affecting other users.

@@

% ---------------------------------------------------------------------------
\subsection{Characterizing the PanDA Broker on Titan}
\label{ssec:broker_titan}

@@

Fig.~\ref{fig:backfill-utilization} shows the total backfill availability on
Titan (gray bars) and the number of core-hours of that availability used by
ATLAS (blue bars) during the experiment time window. ATLAS consumed a total
of 51.4M core-hours, for an average of $\approx$3.7M core-hours a month. 
% , with a minimum of 1.7M core-hours in April 2016 and a maximum 7.9M
% core-hours in February 2017.

\begin{figure}[!t]
    \includegraphics[clip,width=\columnwidth]{figures/backfill_consumption.pdf}
    \vspace{-0.3in}
    \caption{Titan's total backfill availability: CPU core-hours (gray) and
    CPU core-hours used by ATLAS (blue). GPU core-hours unaccounted for as
    they cannot be used by ATLAS\@. Efficiency of PanDA Brokers defined as
    percentage of total Titan's backfill availability used by ATLAS (Red
    labels).}
\label{fig:backfill-utilization}
\end{figure}

@@

Every detector simulation executed on Titan process 100 events. This number
of events is consistent with the physics of the use case and with the average
duration of backfill availability. The duration of a detector simulation is a
function of the number of events simulated but not all events take the same
time to be simulated. One event simulation takes from $\approx$2 to
$\approx$40 minutes, with a mean of $\approx$14 minutes. Considering that
each worker node process up to 16 events concurrently, 100 events takes an
average of 105 minutes to process. As such, PanDA brokers do not use backfill
availability with less than 105 minutes walltime.

@@

%\begin{figure*}[!t]
\begin{figure}[!t]
    % \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_wnodes_distribution.pdf}
    % \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_walltime_distribution.pdf}
    % \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_avail.png}
    \includegraphics[clip,width=\columnwidth]{figures/titan_backfill_avail.png}
   \vspace{-0.1in}
    \caption{
    % Backfill availability: distribution of number of work nodes (left) and
    % of walltime in minutes (center). Scatter plot of the two variables.
    % (right).
    62555 measures of Backfill availability on Titan during the experiment
    time window. Mean number of work nodes available 691; mean walltime
    available 126 minutes.}
\label{fig:backfill-distrib}
%\end{figure*}
\end{figure}

% Usually, detector simulations performed on the Grid process 10 times the
% number of events than those on Titan. This explains the difference between
% the percentage of detector simulations and of events computed by those
% simulations on Titan.
PanDA Broker could fit the number of events to the walltime of each available
backfill slot on the base of the distributions of the time taken by one event
to be simulated. That specific number of event could then be pulled from the
PanDA Event service~\cite{calafiura2015atlas} and given as input to one or
more simulations. Once packaged into the MPI script submitted to titan's PBS
batch system, these simulations would better fit their available backfill
slot, contributing to increase the efficiency of PanDA Brokers.

@@

In September 2016, increasing the number of concurrent PanDA brokers from 4
to 20 markedly improved efficiency (see Fig.~\ref{fig:backfill-utilization})
but further research is ongoing to understand whether an even greater
number of brokers would yield even greater efficiency. This research focuses on
evaluating the overheads of input/output files staging, including its impact
on DTNs, and on an alternative design of PanDA Broker that enables the
concurrent submission of multiple MPI scripts~\cite{barreiro2016panda}. The
understanding will contribute to improving the efficiency of
PanDA Brokers beyond the 30\% limit showed in
Fig.~\ref{fig:backfill-utilization}.

The current design and architecture of the PanDA Broker is proving to be as
reliable as PanDA Pilot when used on the WLCG\@. Between Jan 2016 and Feb
2017, the overall failure rate of all the ATLAS detector simulation jobs was
14\%, while the failure rate of jobs submitted to Titan was a comparable
13.6\%. PanDA Brokers were responsible for around the 19\% of the failures,
compared to the 29\% of failures produced by the JobDispatcher module of the
PanDA Server, and the 13\% failures produced by the Geant4 toolkit. 
% The current failure rate of the PanDA Brokers confirms the benefits of
% reusing most of the code base of the PanDA Pilot for implementing the PanDA
% Broker. It also shows that adopting third-party libraries like RADICAL-SAGA
% did not have a measurable adverse effect on reliability.

% ---------------------------------------------------------------------------
\subsection{Characterizing the Detector Simulation on Titan}
\label{ssec:athenamp_titan}

@@

AthenaMP has an initialization and configuration stage. At initialization
time, AthenaMP is assembled from a set of shared libraries, depending on the
type of payload that will have to be computed. Once initialized, every
algorithm and service of AthenaMP is configured via Python scripts. Both
these operations result in read operations on the filesystem shared between
the worker nodes and the DTNs, including the operations required to access
small python scripts.

Initially, all the shared libraries and the python scripts of AthenaMP were
stored on the OLCF Spider 2 Lustre file system. However, the I/O patterns of
the initialization and configuration stages degraded the performance of the
filesystem. This was addressed by moving the AthenaMP distribution to a
read-only NFS directory, shared among Titan's DTNs and worker nodes. NFS
eliminated the problem of metadata contention, improving metadata read
performance from $\approx$6,300 seconds on Lustre to $\approx$1,500 seconds
on NFS\@.

@@

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx8_tx16_comparison_vsquashed.pdf}
   \vspace{-0.3in}
    \caption{Distributions of the time taken to simulate one event when
    placing 2 simulations (h1) or 1 simulation (h2) per Titan's CPU\@. 2
    simulation use 16 cores per node, 1 simulation 8. 50 Events; 1 Titan
    worker nodes; 16 work threads per node; 100 events per node.}

\label{fig:comparison-8-16cores}
\end{figure}

The performance analysis of Titan's AMD CPUs for detector simulations helps
also to compare Titan and Grid site performances. Usually, Grid sites exposes
resources with heterogeneous CPU architectures and a maximum of 8 (virtual)
cores per worker node, while Titan's offer an homogeneous 16 cores
architecture. We used the rate of events processes per minute as a measure of
the efficiency of executing the same detector simulation on Titan or Grid
sites. Comparisons of the efficiencies of Titan to the BNL and SIGNET Grid
sites, normalized for 8 cores, show that the effective performance per-core
at Titan is $\approx$0.57 event per minute, roughly 1/2 of BNL and  1/3 of
SIGNET performances.

@@

\begin{table*}[t]
\centering
\begin{tabular}{lllllllll} & Num. Nodes & Duration (s) & Read (GB) & Written
 (GB) & GB Read/nodes & GB Written/nodes & $open()$ & $close()$ \\ Min & 1 &
 1,932 & 0.01 & 0.03 & 0.00037 & 0.02485 & 1,368 & 349 \\ Max & 300 & 7,452 &
 241.06 & 71.71 & 0.81670 & 0.23903 & 1,260,185 & 294,908 \\ Average & 35.66
 & 6,280.82 & 20.36 & 6.87 & 0.38354 & 0.16794 & 146,459.37 & 34,155.74 \\
 Std. Dev. & 55.33 & 520.99 & 43.90 & 12.33 & 0.19379 & 0.03376 & 231,346.55
 & 53,799.08
\end{tabular}
\caption{The Statistical breakdown of the I/O impact of 1,175 jobs ATLAS
executed at OLCF for the week of 10/25/16}
\label{panda-olcf-stats}
\end{table*}

@@

The metadata I/O breakdown shows that ATLAS jobs yield 23 file $open()$
operations per second (not including file $stat()$ operations) and 5 file
$close()$ operations per second, with similar distributions. On average, the
maximum number of file $open()$ operations per job is $\approx$170/s and the
maximum number of file $close()$ operations is $\approx$39/s. For the 1,175
ATLAS jobs observed, the total number of file $open()$ operations is
172,089,760 and the total number of file $close()$ operations is 40,132,992.
The difference between these two values is still under investigation: One
possible explanation is that ATLAS jobs don't call a file $close()$ operation
per every file $open()$ issued.

>>>>

\ifreview
Comments to address in this section:
\begin{enumerate}
    \color{red}

    \item REVIEWER 2: What are the reasons for the low efficiency of PanDA
    Broker? Why are only 30\% of backfill slots useable, if they are
    basically uncontested?

    \item REVIEWER 2: Why does Fig 3 show an increase of absolute and
    relative efficiency in Aug-16, when the text claims this in Sep-16 due to
    more PanDA Brokers?\mtnote{Fixed.}

    \item REVIEWER 2: Why are only 20 PanDA Brokers deployed if this is a
    limiting factor? Why is it questionable that more PanDA Brokers would
    increase efficiency? If only 5/6 of resources below the mean are
    unusable, this implies an efficiency of almost 60\% being
    reachable.\mtnote{From our text: ``further research is ongoing to
    understand whether an even greater number of brokers would yield even
    greater efficiency. This research focuses on evaluating the overheads of
    input/output files staging, including its impact on DTNs, and on an
    alternative design of PanDA Broker that enables the concurrent submission
    of multiple MPI scripts''. I believe this is an answer good enough.}

    \item REVIEWER 2: Given an expected increase in resource need of 10-100,
    how is the capacity of 4\% WLCG significant? How does this relate to the
    manpower required to develop and maintain a custom solution per HPC site?
    How many TITAN resources are actually available to HEP, given the use by
    other domains? The magnitude of capacities presented as available is in
    major contrast to the requirements sketched in the introduction.

    \item REVIEWER 2: How does the poor metadata read performance of
    1500-6300s (25\%-100\% of minimum walltime!) affect efficiency? Is it
    factored into the choice of walltime requests?\mtnote{Asked Sergey and
    Danila to confirm these numbers and walltime request policy.} Is it
    applicable only to TITAN, to HPC in general, or even to regular WLCG
    sites?\mtnote{This is clearly out of scope for this paper.} Are there
    solutions at application, e.g. the use of packaged python scripts and
    libraries?\mtnote{Added the following: ``While these figures are specific
    to OLCF and TItan, reengineering on AthenaMP could further improve
    startup time on every platform.''} The information presented is
    insufficient to estimate whether this issue applies to other environments
    and usecases as well, or is simply due to suboptimal
    implementation/configuration.

    \item REVIEWER 3: The HEP-SPEC06 benchmark, which is used as a
    ``computing power'' measure in LCG is mostly integer-artihmetic based: FP
    operations accounts for \~10\% of the average workload (see [1], section
    ``Why do you use `all\_cpp' instead of SPECint or SPECfp?'') While this
    figure certainly comes from averaging over many different workloads
    across all experiments, it seems a bit surprising that FP scheduling
    contention is responsible for 50\% decrease in compute efficiency on
    Titan when compared to other WLCG sites (paper page 6).\mtnote{Wrote an email to Sergey asking for a clarification.}
\end{enumerate}
\fi

@@

@@

% ---------------------------------------------------------------------------
\subsection{Characterizing the PanDA Broker on Titan}\label{ssec:broker_titan}

@@

Fig.~\ref{fig:backfill-utilization} shows the total backfill availability on
Titan (gray bars) and the number of core-hours of that availability used by
ATLAS (blue bars) during the experiment time window. ATLAS consumed a total
of 51.4M core-hours, for an average of \(\approx\)3.7M core-hours a month.

\begin{figure}[!t]
    \includegraphics[clip,width=\columnwidth]{figures/backfill_consumption.pdf}
    \vspace{-0.3in}
    \caption{Titan's total backfill availability: CPU core-hours (gray) and
    CPU core-hours used by ATLAS (blue). GPU core-hours unaccounted for as
    they cannot be used by ATLAS\@. Efficiency of PanDA Brokers defined as
    percentage of total Titan's backfill availability used by ATLAS (Red
    labels).}\label{fig:backfill-utilization}
\end{figure}

@@

Every detector simulation executed on Titan process 100 events. This number
of events is consistent with the physics of the use case and with the average
duration of backfill availability. The duration of a detector simulation is a
function of the number of events simulated but not all events take the same
time to be simulated. One event simulation takes from \(\approx\)2 to
\(\approx\)40 minutes, with a mean of \(\approx\)14 minutes. Considering that
each worker node process up to 16 events concurrently, 100 events takes an
average of 105 minutes to process. As such, PanDA brokers do not use backfill
availability with less than 105 minutes walltime.

@@

\begin{figure}[!t]
    \includegraphics[clip,width=\columnwidth]{figures/titan_backfill_avail.png}
    \vspace{-0.1in}
    \caption{62555 measures of Backfill availability on Titan during the
    experiment time window. Mean number of work nodes available 691; mean
    walltime available 126 minutes.}\label{fig:backfill-distrib}
\end{figure}

PanDA Broker could fit the number of events to the walltime of each available
backfill slot on the base of the distributions of the time taken by one event
to be simulated. That specific number of event could then be pulled from the
PanDA Event service~\cite{calafiura2015atlas} and given as input to one or
more simulations. Once packaged into the MPI script submitted to titan's PBS
batch system, these simulations would better fit their available backfill
slot, contributing to increase the efficiency of PanDA Brokers.

@@

In August 2016, increasing the number of concurrent PanDA brokers from 4 to
20 markedly improved efficiency (see Fig.~\ref{fig:backfill-utilization}) but
further research is ongoing to understand whether an even greater number of
brokers would yield even greater efficiency. This research focuses on
evaluating the overheads of input/output files staging, including its impact
on DTNs, and on an alternative design of PanDA Broker that enables the
concurrent submission of multiple MPI scripts~\cite{barreiro2016panda}. The
understanding will contribute to improving the efficiency of PanDA Brokers
beyond the 30\% limit showed in Fig.~\ref{fig:backfill-utilization}.

The current design and architecture of the PanDA Broker is proving to be as
reliable as PanDA Pilot when used on the WLCG\@. Between Jan 2016 and Feb
2017, the overall failure rate of all the ATLAS detector simulation jobs was
14\%, while the failure rate of jobs submitted to Titan was a comparable
13.6\%. PanDA Brokers were responsible for around the 19\% of the failures,
compared to the 29\% of failures produced by the JobDispatcher module of the
PanDA Server, and the 13\% failures produced by the Geant4 toolkit.

% ---------------------------------------------------------------------------
\subsection{Characterizing the Detector Simulation on
Titan}\label{ssec:athenamp_titan}

@@

AthenaMP has an initialization and configuration stage. At initialization
time, AthenaMP is assembled from a very large set of shared libraries,
depending on the type of payload that will have to be computed. Once
initialized, every algorithm and service of AthenaMP is configured via Python
scripts. Both these operations result in read operations on the filesystem
shared between the worker nodes and the DTNs, including the operations
required to access small python scripts.

Initially, all the shared libraries and the python scripts of AthenaMP were
stored on the OLCF Spider 2 Lustre file system. However, the I/O patterns of
the initialization and configuration stages degraded the performance of the
filesystem. This was addressed by moving the AthenaMP distribution to a
read-only NFS directory, shared among Titan's DTNs and worker nodes. NFS
eliminated the problem of metadata contention, improving metadata read
performance from \(\approx\)6,300 seconds on Lustre to \(\approx\)1,500
seconds on NFS\@. While these figures are specific to OLCF and TItan, reengineering on AthenaMP could further improve startup time on every platform.

@@

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx8_tx16_comparison_vsquashed.pdf}
    \vspace{-0.3in}
    \caption{Distributions of the time taken to simulate one event when
    placing 2 simulations (h1) or 1 simulation (h2) per Titan's CPU\@. 2
    simulation use 16 cores per node, 1 simulation 8. 50 Events; 1 Titan
    worker nodes; 16 work threads per node; 100 events per
    node.}\label{fig:comparison-8-16cores}
\end{figure}

The performance analysis of Titan's AMD CPUs for detector simulations helps
also to compare Titan and Grid site performances. Usually, Grid sites exposes
resources with heterogeneous CPU architectures and 8 (virtual) cores, while
Titan's offer an homogeneous 16 cores architecture. We used the rate of
events processes per minute as a measure of the efficiency of executing the
same detector simulation on Titan or Grid sites. Comparisons of the
efficiencies of Titan to the BNL and SIGNET Grid sites, normalized for 8
cores, show that the effective performance per-core at Titan is
\(\approx\)0.57 event per minute, roughly 1/2 of BNL and  1/3 of SIGNET
performances.

@@

\begin{table*}[t]
\centering
\begin{tabular}{lllllllll} & Num. Nodes & Duration (s) & Read (GB) & Written
 (GB) & GB Read/nodes & GB Written/nodes & \(open()\) & \(close()\) \\ Min & 1 &
 1,932 & 0.01 & 0.03 & 0.00037 & 0.02485 & 1,368 & 349 \\ Max & 300 & 7,452 &
 241.06 & 71.71 & 0.81670 & 0.23903 & 1,260,185 & 294,908 \\ Average & 35.66
 & 6,280.82 & 20.36 & 6.87 & 0.38354 & 0.16794 & 146,459.37 & 34,155.74 \\
 Std. Dev. & 55.33 & 520.99 & 43.90 & 12.33 & 0.19379 & 0.03376 & 231,346.55
 & 53,799.08
\end{tabular}
\caption{The Statistical breakdown of the I/O impact of 1,175 jobs ATLAS
executed at OLCF for the week of 10/25/16}\label{panda-olcf-stats}
\end{table*}

@@

The metadata I/O breakdown shows that ATLAS jobs yield 23 file \(open()\)
operations per second (not including file \(stat()\) operations) and 5 file
\(close()\) operations per second, with similar distributions. On average,
the maximum number of file \(open()\) operations per job is \(\approx\)170/s
and the maximum number of file \(close()\) operations is \(\approx\)39/s. For
the 1,175 ATLAS jobs observed, the total number of file \(open()\) operations
is 172,089,760 and the total number of file \(close()\) operations is
40,132,992. The difference between these two values is still under
investigation: One possible explanation is that ATLAS jobs don't call a file
\(close()\) operation per every file \(open()\) issued.
